The Error Estimate for Computing Electrostatic Interaction in the Inhomogeneous and Correlated Molecular Systems: for the Ewald Summation, SPME and Staggered Mesh Ewald.

### 2012-03-23 longRange.hetero/publish/main.tex
nearly finished, ask for advice

The calculation of long-range electrostatic interaction is a very
important topic in the molecular simulations.  Due to the very slow
decay of the interaction with respect to the distance between charges,
the cut-off and neighbor list algoritm does not work, and moreover the
energy of the system is only conditionally convergent.  One of the
first algoritms handling this problem is the Ewald summation that
dates back to 1920's~\cite{ewald1921die}.  However, it is shown that
with a careful tuning of working parameters, the optimal computational
cost of the Ewald summation is of $\mathcal
O(N^{3/2})$~\cite{perram1988asc}, which is not feasible for morden
large scale molecular simulations. The particle mesh Ewald (PME)
method~\cite{darden1993pme}, the smooth particle mesh Ewald (SPME)
method~\cite{essmann1995spm} and the particle-particle-particle-mesh
(P3M) method~\cite{hockney1988computer, deserno1998mue1} are the most
popular Ewald-based fast algorithms that reduces the computational
cost to $\mathcal O(N\log N)$. More recently, the staggered mesh or
``interlacing'' technique firtly developed in
1970's~\cite{chen1974reduction,eastwood1976optimal} is applied to SPME
(called staggered mesh Ewald)~\cite{cerutti2009staggered} and P3M
(called interlaced P3M)~\cite{neelov2010interlaced}, and is showed to
improve the accuracy of these algorithms.

The fast algoritms have achieved great success in the computer
simulation of charged molecular systems over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is diffcult to try out the
most efficient parameter set, and an arbitary choice may lead to
qualitively wrong simulation result. This problem can be solved by an
in-depth study of the error introduced by these algrotims, namly the
error estimate~\cite{petersen1995accuracy, deserno1998mue2,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameter.  By these error estimates, a work flow was designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is fastest one given a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of them is satisfactorily fulfilled in most
realistic systems of practical interest.  For example, the lipid biomembrane
solving in water: the carbon chains of the lipids carry no charge, and
the charges are correlated by covalent bonds, van der Waals
interaction, hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  A lot of
research have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

The purpose of this research is to develop an realiable force error
estimate in inhomogeneous and strongly correlated molecular systems.
We estimate the force error because the application is the molecular
dynamics (MD) simulation.  Firstly, a most general error estimate
frame work for the pairwise interactions is set up. We prove that the
error is decomposed into three additive parts: the homogeneity error,
the inhomogeneity error and the correlation error~\cite{wang2012}. The
homogeneity and inhomogeneity error can be calculated at a cost of
$\mathcal O(N\log N)$ given the convolutional form of the error
estimates.  Then a way is proposed to calculate the approximation of
correlation error at cost of $\mathcal O(N\log N)$. Though it is
possible to improve the accuracy of the approximation systematically,
numerical results show that it is good enough
to consider the correlation stems from
the neighboring atoms within the same molecule
in an inhomogeneous water system.  Following this framework, we
develop the error estimate for the Ewald summation, the SPME method
and the staggered method Ewald method. We study the SPME method
because it is the one of the most popular long-range algorithms among
the mainstream MD simulation packages, for example,
AMBER~\cite{case2005amber}, GROMACS~\cite{van2005gromacs,
  hess2008gromacs} and NAMD~\cite{phillips2005scalable}.  The estimate
of the homogeneity and inhomogeneous errors is tested by by two
randomly distributed inhomogeneous charges systems, and is proved to
be very sharp in these ideal systems.  By studying a real
inhomogeneous water system, the nearest neigbor approximation of
correlation error is proved to improve the quality of error estimate
greatly. This paper is ended by conclusions and remarks in
section~\ref{sec:conclusions}.

@@

In this paper, we consider the force error, bacause the applicaions
are based on the molecular dynamics simulation.  Suppose the system is
composed by $N$ charged particles located at $\v r_1, \v r_2, \cdots,
\v r_N$ with charges $q_1, q_2, \cdots, q_N$, respectively.  We study
the force error of an arbitrary charge $i$ by calculating the
difference between the exact force and calculated force exerting on
the charge $i$, which is called the \emph{error force}
\cite{wang2012} and denoted by $\Delta \v F(\v r_i)$.
% and is denoted by
% $\Delta \v F(\v r_i) = \v F(\v r_i) - \tilde{\v F}(\v r_i)$.
The maganitude of the force computation error 
% error of the force computation
is defined by the \emph{root mean square (RMS)} \emph{error}, which is
the square root of the second moment of the error force: $\mathcal
E(\v r_i) = \sqrt{\langle\vert\Delta\v F(\v r_i)\vert^2\rangle}$.  The
notation $\langle\cdot\rangle$ means ensemble average.  We also
estimate the the ensemble average of the error force (also called
\emph{mean error force}), namly $\langle\Delta\v F(\v r_i)\rangle$,
which plays an important role in the error analysis of the
short-range inhomogeneous systems~\cite{wang2012}.

@@

  If the error force of an arbitary particle $i$ is

@@

@@

    \d d\v r'\d d\v r''.

@@

  Where $\ast$ denotes the convolution, and

@@

  then the error estimates are in the convolution form, namely:

@@

  Where $\ast$ denotes the convolution.

@@

  particles in the system~\cite{kolafa1992cutoff, deserno1998mue2, wang2010optimizing, wang2012},

@@

  By writting the error estimates in the covolution form in

@@

\subsection{Nearest neighbor approximation of the correlation
error}

In this section, we consider the nearest neighbor approximation of the
correlation error. Here the definition of the ``nearest neighbor'' is
very flexible, basically it means the neigbors of a charge that are
not very far away. It could be defined by the charges falling in some
certain neighboring range or charge connected by a chemical bonds.
% Here ``first neighbors'' means the neighboring
% atoms that are connected by the chemical bond.  In most molecular
% force fields, such bonds are modeled by a rigid connection between two
% atoms.  Therefore, the correlation between such an atom pair is very
% strong.
% In the system of water, the relative position of the oxygen and
% hydrogen atoms within one molecule is fixed. Therefore, according
% to this property, it is possible to improve the quality of the
% error estimate.
Starts from Eqn.~\eqref{eqn:prf-error-step3},
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r_i)\vert^2\rangle
  =\,&
  q_i^2\int_{\mathbb R^3}
  \vert\v K(\v r_i , \v r')\vert^2\rho_{q^2}(\v r')\,\d d\v r'
  \\
  & +
  q_i^2\int_{\mathbb R^3\times\mathbb R^3}
  \v K(\v r_i , \v r')\cdot\v K(\v r_i , \v r'')\,
  \rho^{(2)}(\v r', \v r'')
  \,\d d\v r'\d d\v r'',
\end{align}
where the definition of $\rho^{(2)}(\v r', \v r'')$ is given by
Eqn.~\eqref{eqn:def-rho2}.
% \begin{align}
%   q_{q^2}(\v r', \v r'') =
%   \bigg\langle
%   \sum_{j\neq k}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
%   \bigg\rangle.
% \end{align}
We seperately consider the  contribution from nearest neigbors
and other contributions to
the pairwise density $\rho^{(2)}(\v r', \v r'')$.  Denoting
$\Omega_j = \{ \,k\,\vert\,
\textrm{atom $k$ is the nearest neighbor of atom $j$} \,\}$,
% the set of the first neighbors of the $j$-th atom by $\Omega_j$,
we have:
\begin{align}\label{eqn:split-rho-q2}
  \rho^{(2)}(\v r', \v r'')
  = &\,
  \bigg\langle
  \sum_j\sum_{k\in\Omega_j}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle
  +
  \bigg\langle
  \sum_j\sum_{k\notin\Omega_j}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle.
\end{align}
We assume two charges are independent if they are not nearest neighbors.
Then
second part on the r.h.s. of the Eqn.~\eqref{eqn:split-rho-q2} is
approximately $\frac{N-\#(\Omega_j)}{N}\rho_q(\v r')\rho_q(\v r'')\approx
\rho_q(\v r')\rho_q(\v r'')$, where $\#(\Omega_j)$ is the number of nearest neighbors considered. This approximation is valid only when the number
of neigbors considered is small comparing with the number of charges
in the system.
% If the number of atoms in the system is
% large, this should be a good approximation.
Therefore, the first part
on the r.h.s. of Eqn.~\eqref{eqn:split-rho-q2} is actually the
pairwise charge corrlation function:
\begin{align}\label{eqn:fnc-correlation-water}
  C_{q^2}(\v r', \v r'') \approx
  \bigg\langle
  \sum_j\sum_{k\in\Omega_j}q_jq_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle.
\end{align}
% If connected by the chemical bond, the position of the first neighbors
% $k$ is not independent with the atom $j$, so their correlation is
% strong, and should be calculated explicitly to improve the quality of
% the error estimate.
By inserting Eqn.~\eqref{eqn:fnc-correlation-water} into
the correlation error, we have:
\begin{align}
  \mathcal E_{\textrm{correlation}}(\v r_i)
  \approx q_i^2\sum_j\sum_{k\in\Omega_j}q_jq_k
  \big\langle
  \v K(\v r_i , \v r_j)\cdot\v K(\v r_i , \v r_k)
  \big\rangle_{\v r_j,\v r_k}
\end{align}
The quality of this approximation depends on how to define the
``nearest neighbors''. In the system of liquid and gas, the
correlation is only short-ranged, so it is possible to systematically
impove the accuracy by considering more and more neigbors as long as
the number of neigbors is small comparing with the size of the system.

As an example, we consider a typical three point charge water model
TIP3P~\cite{jorgensen1983comparison}: each hydrogen atom carries a
charge of $q_{\textrm{H}} = +0.417e$ and each oxygen atom carries a
charge of $q_{\textrm{O}} = -0.834e$. Within one molecule, the H-O
bond and H-O-H angle are rigid, so the correlation within one molecule
is very strong.  Therefore, in this case, we define the nearest neighbor by the
atoms within the same water molecule.  Denoting the index set of the oxygen
atoms by $\Omega_{\textrm{O}}$, and the index set of the hydrogen
atoms by $\Omega_{\textrm{H}}$, we have:
\begin{align}\nonumber
  \mathcal E_{\textrm{correlation}}(\v r_i)
  = &\,
  q_i^2\sum_{j\in\Omega_{\textrm{O}}}\sum_{k\in\Omega_j}q_jq_k
  \big\langle
  \v K(\v r_i , \v r_j)\cdot\v K(\v r_i , \v r_k)
  \big\rangle_{\v r_j,\v r_k} \\ \nonumber
  &\,+ 
  q_i^2\sum_{j\in\Omega_{\textrm{H}}}\sum_{k\in\Omega_j}q_jq_k
  \big\langle
  \v K(\v r_i , \v r_j)\cdot\v K(\v r_i , \v r_k)
  \big\rangle_{\v r_j,\v r_k} \\\nonumber
  = &\,
  q_i^2
  \int_{\mathbb R^3}
  \v K(\v r_i , \v r')\cdot
  2\big\langle
  q_{\textrm{H}}\v K(\v r_i , \v r' + \v s_{\textrm{O}})
  \big\rangle\,
  \rho_{\textrm{O}}(\v r')
  \d d\v r' \\\label{eqn:c-error-1}
  & \,+
  q_i^2
  \int_{\mathbb R^3}
  \v K(\v r_i , \v r')\cdot
  \big\langle
  q_{\textrm{O}}\v K(\v r_i , \v r' - \v s_{\textrm{O}})+
  q_{\textrm{H}}\v K(\v r_i , \v r' + \v s_{\textrm{H}})
  \big\rangle\,\rho_{\textrm{H}}(\v r')
  \d d\v r'
\end{align}
If positions of the three atom within one molecule are
$\v r_{\textrm{H}_1}$, $\v r_{\textrm{H}_2}$ and $\v r_{\textrm{O}}$, then
% $\v s_{\textrm{H}_1} = \v r_{\textrm{H}_1} - \v r_{\textrm{O}}$, 
% $\v s_{\textrm{H}_2} = \v r_{\textrm{H}_2} - \v r_{\textrm{O}}$,
% $\v s_{\textrm{O}} = \v r_{\textrm{O}} - \v r_{\textrm{H}_1}$ and
% $\v s_{\textrm{H}} = \v r_{\textrm{H}_2} - \v r_{\textrm{H}_1}$.
$\v s_{\textrm{O}} = \v r_{\textrm{H}_1} - \v r_{\textrm{O}}$ and
$\v s_{\textrm{H}} = \v r_{\textrm{H}_2} - \v r_{\textrm{H}_1}$.
The two ensemble averages in Eqn.~\eqref{eqn:c-error-1}
is taken over all possible water directions 
at position $\v r'$. As an approximation, we
assume these averages are  spacially uniform, namely independet with $\v r'$.
This assumption may be  problematic
in the regions where the preference of water  drection is different
with others.

To calculate the nearest neighbor approximated correlation error,
we assume the error force kernel
has the form $\v K(\v r_i , \v r_j) = \v K(\v r_i - \v r_j)$, so
the integrals can be written in the covolution form and accelerated by the FFT.
% In the example considered by the present paper in section~\ref{sec:tmp3},
% it is shown to be a good approximation to obtain the estimate of the
% correlation error.
One possible way to calculate the ensemble averages in
Eqn.~\eqref{eqn:c-error-1} is by the Fourier transform
of the error force kernel:
\begin{align} \nonumber
  \big\langle
  q_{\textrm{H}}\v K(\v r_i - \v r' + \v s_{\textrm{O}})
  \big\rangle 
  =\,&
  \frac1V\bigg\langle
  \sum_{\v m}
  q_{\textrm{H}}\hat{\v K}(\v m)
  e^{2\pi i\v m\cdot(\v r_i - \v r' + \v s_{\textrm{O}})} 
  \bigg\rangle \\\nonumber
  =\,&
  \frac1V\sum_{\v m}
  \langle
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{O}}}
  \rangle\,
  \hat{\v K}(\v m)\,
  e^{2\pi i\v m\cdot(\v r_i - \v r')},
\end{align}
where ``$\wedge$'' denotes the forward Fourier transform. By denoting
\begin{align}
  \hat T_{\textrm{O}}(\v m)
  &= 
  2\langle
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{O}}}
  \rangle,\\
  \hat T_{\textrm{H}}(\v m)
  &= 
  \langle
  q_{\textrm{O}}\,e^{-2\pi i\v m\cdot\v s_{\textrm{O}}} +
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{H}}}
  \rangle.
\end{align}
The Eqn.~\eqref{eqn:c-error-1} becomes:
\begin{align}\nonumber
  \mathcal E_{\textrm{correlation}}(\v r_i)
  =&\,
  q_i^2
  \int_{\mathbb R^3}
  \v K(\v r_i - \v r')\cdot
  (\hat T_{\textrm{O}}\hat{\v K})^{\vee}(\v r_i - \v r')
  \,\rho_{\textrm{O}}(\v r')
  \,\d d\v r' \\\label{eqn:c-error-water}
  & \,+
  q_i^2
  \int_{\mathbb R^3}
  \v K(\v r_i - \v r')\cdot
  (\hat T_{\textrm{H}}\hat{\v K})^{\vee}(\v r_i - \v r')
  \,\rho_{\textrm{H}}(\v r')
  \,\d d\v r',
\end{align}
where ``$\vee$'' denotes the backward Fourier transform.
In the Fourier space, the error force kernel $\v K(\v r)$ is
influenced by the prefactors $T_{\textrm{O}}$ and $T_{\textrm{H}}$,
which are accounting for the atomic correlation within one water
molecule.

@@

From the precise calculation of the reciprocal summation to the SPME
fast algorithm, the electrostatic interaction is approximated by two
steps.  The first step is the truncation of the reciprocal summation,
i.e. Eqn. \eqref{eqn:tmp6}. The second step is the approximation of
$e^{2\pi i\v m\cdot\v r}$ by B-spline
interpolation. Ref. \cite{wang2010optimizing} has pointed out that the
error due to the first step is negligible comparing with the second
step, at least for the parameters of practical interest. Numerical
examples supporting this argument in inhomogeneous systems are also
given in Section \ref{sec:tmp3}.  Therefore, to estimate the force
error, it is enough to compare the SPME force with the
\emph{truncated} Ewald force \eqref{eqn:tmp8}.

As mentioned above, the SPME error mainly oringinates from approximating

@@

the analytical expression of $A(\v m, \v r)$:

@@

The error force of the ik-differentiation is therfore

@@

estimates, and leave all details of calculating the error estimates

@@

valleys of the high-wave-number terms may canncels in the the
emsemble-averaged error estimate.

@@

so it can be calculated once and stored in a table for further uses.

@@

\subsection{Error estimate of the analytical differetiation}
\label{sec:error-ana}

>>>>

The calculation of the long-range electrostatic interaction is a very
important topic in the molecular simulations.  Due to the very slow
decay of the interaction with respect to the distance between charges,
the short-range fast algorithms (the cell list and neighbor list
algorithms) do not work, and the
energy of the system is only conditionally convergent.  One of the
first algorithms handling this problem was the Ewald summation, dating
back to the 1920's~\cite{ewald1921die}.  However, it has been shown that
with a careful tuning of working parameters, the optimal computational
cost of the Ewald summation is of $\mathcal
O(N^{3/2})$~\cite{perram1988asc}, which is not feasible for modern
large scale molecular simulations.
Therefore, several Ewald based fast algorithms were proposed to reduce the
computational cost to a scalable level: $\mathcal O(N\log N)$. Some of them
are
the particle mesh Ewald (PME)
method~\cite{darden1993pme}, the smooth particle mesh Ewald (SPME)
method~\cite{essmann1995spm} and the particle-particle-particle-mesh
(P3M) method~\cite{hockney1988computer, deserno1998mue1}.
More recently, the staggered mesh or
``interlacing'' technique~\cite{chen1974reduction,eastwood1976optimal}
was applied to SPME
(called staggered mesh Ewald)~\cite{cerutti2009staggered} and P3M
(called interlaced P3M)~\cite{neelov2010interlaced}, and was showed to
improve the accuracy of these algorithms.
In this paper, we focus on the Ewald summation, the SPME and
the staggered mesh Ewald methods. The Ewald summation is the starting
point of all the mentioned fast algorithm. SPME
is the one of the most popular long-range algorithms among
the mainstream molecular simulation packages, for example,
AMBER~\cite{case2005amber}, GROMACS~\cite{van2005gromacs,
  hess2008gromacs} and NAMD~\cite{phillips2005scalable}.

The aforementioned fast algorithms have achieved great success
over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is difficult to try out the
most efficient parameter set in such a high dimensional parameter space.
The experiential choices may be either less accurate or waste
computational effort.
This problem can be solved by an
in-depth study of the error introduced by these algorithms, namely the
error estimate~\cite{kolafa1992cutoff, hummer1995numerical,
  petersen1995accuracy, deserno1998mue2,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameters.  Using the error estimates, a work flow has been designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is the fastest one given a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of the assumptions is satisfactorily
fulfilled in most molecular
systems of practical interest.  For example, the lipid biomembrane
solving in water: the carbon chains of the lipids carry no charge, and
the charges are correlated by the covalent bonds, the van der Waals
interaction, the hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  A lot of
research have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

The purpose of this research is to develop a reliable  error
estimate in inhomogeneous and correlated molecular systems.
We estimate the force error because the application is the molecular
dynamics (MD) simulation.  Firstly, a general error estimate
frame work for the pairwise interaction is set up. We prove that the
error is decomposed into three additive parts: the homogeneity error,
the inhomogeneity error and the correlation error~\cite{wang2012}.
These error estimates are only of theoretical interest, and
not applicable, because the computational costs of the error estimates
are $\mathcal O(N^2\log N)$ (homogeneity and inhomogeneity error)
and $\mathcal O(N^3\log N)$ (correlation error).
If the error force has a special form so that the error estimates
can be written as convolutions, the computational costs
are reduced to $\mathcal O(N\log N)$ and 
$\mathcal O(N^2\log N)$, respectively.
The computational cost of the correlation error is further
reduce to $\mathcal O(N\log N)$ by the nearest neighbor approximation
technique, which assume the system is only short-range correlated. This
is the case for most the liquid and gas system far from critical point.
Following this framework, we
develop the error estimates for the Ewald summation, the SPME method
and the staggered method Ewald method.   The estimates
of the homogeneity and inhomogeneous errors are tested by two
randomly distributed inhomogeneous charges systems
(i.e. the charges are not correlated), and is proved to
be very sharp.
The error estimates are also tested in a real inhomogeneous water system,
in which the charge correlation plays a very important role
on the force error. By 
including the nearest neighbor approximation of
correlation error, the quality of the estimates
is impressively enhanced.
This paper is ended by conclusions and remarks in
section~\ref{sec:conclusions}.

% The
% homogeneity and inhomogeneity error can be calculated at a cost of
% $\mathcal O(N\log N)$ given the convolutional form of the error
% estimates.  Then a way is proposed to calculate the approximation of
% correlation error at cost of $\mathcal O(N\log N)$. Though it is
% possible to improve the accuracy of the approximation systematically,
% numerical results show that it is good enough
% to consider the correlation stems from
% the neighboring atoms within the same molecule
% in an inhomogeneous water system.

% Following this framework, we
% develop the error estimate for the Ewald summation, the SPME method
% and the staggered method Ewald method.   The estimate
% of the homogeneity and inhomogeneous errors is tested by by two
% randomly distributed inhomogeneous charges systems, and is proved to
% be very sharp in these ideal systems.  By studying a real
% inhomogeneous water system, the nearest neigbor approximation of
% correlation error is proved to improve the quality of error estimate
% greatly. This paper is ended by conclusions and remarks in
% section~\ref{sec:conclusions}.

@@

In this paper, we consider the system with periodic boundary condition.
Suppose it is
composed by $N$ charged particles located at $\v r_1, \v r_2, \cdots,
\v r_N$ with charges $q_1, q_2, \cdots, q_N$, respectively.  We study
the force error of an arbitrary charge $i$ by calculating the
difference between the exact force and calculated force exerting on
the charge $i$, which is called the \emph{error force}
\cite{wang2012} and denoted by $\Delta \v F(\v r_i)$.
% and is denoted by
% $\Delta \v F(\v r_i) = \v F(\v r_i) - \tilde{\v F}(\v r_i)$.
The magnitude of the force computation error 
% error of the force computation
is defined by the \emph{root mean square (RMS)} \emph{error}, which is
the square root of the second moment of the error force: $\mathcal
E(\v r_i) = \sqrt{\langle\vert\Delta\v F(\v r_i)\vert^2\rangle}$.  The
notation $\langle\cdot\rangle$ means ensemble average.  We also
estimate the the ensemble average of the error force (also called
\emph{mean error force}), namely $\langle\Delta\v F(\v r_i)\rangle$,
which plays an important role in the error analysis of the
inhomogeneous systems~\cite{wang2012}.

@@

  If the error force of an arbitrary particle $i$ is

@@

    \mathcal E^2 (\v r_i) 

@@

    \d d\v r'\d d\v r'',

@@

  where

@@

  then the error estimates are convolutions, namely:

@@

  where ``$\ast$'' denotes the convolution.

@@

  particles in the system~\cite{kolafa1992cutoff, hummer1995numerical, deserno1998mue2, petersen1995accuracy, wang2010optimizing, neelov2010interlaced, ballenegger2012convert, wang2012},

@@

  By writing the error estimates in the convolution form in

@@

\subsection{Nearest-neighbor approximation of the correlation
error}
The full computation of the correlation error cost at least $\mathcal
O(N^2\log N)$, which is still not applicable.
So we consider the nearest-neighbor approximation to reduce
the computational cost.
Here the definition of the ``nearest neighbors'' is
very flexible, basically it means the nearby neighbors that are
strongly correlated. It could also be defined by the charges falling in some
certain neighboring range or charges connected by a chemical bonds.
% Here ``first neighbors'' means the neighboring
% atoms that are connected by the chemical bond.  In most molecular
% force fields, such bonds are modeled by a rigid connection between two
% atoms.  Therefore, the correlation between such an atom pair is very
% strong.
% In the system of water, the relative position of the oxygen and
% hydrogen atoms within one molecule is fixed. Therefore, according
% to this property, it is possible to improve the quality of the
% error estimate.
Starting from Eqn.~\eqref{eqn:prf-error-step3},
  \\
  \,\d d\v r'\d d\v r'',
where the definition of $\rho^{(2)}(\v r', \v r'')$ is given by
Eqn.~\eqref{eqn:def-rho2}.
% \begin{align}
%   q_{q^2}(\v r', \v r'') =
%   \bigg\langle
%   \sum_{j\neq k}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
%   \bigg\rangle.
% \end{align}
We separately consider the  contribution from the nearest neighbors
and other contributions to
the pairwise density $\rho^{(2)}(\v r', \v r'')$.  Denoting
$\Omega_j = \{ \,k\,\vert\,
\textrm{atom $k$ is the nearest neighbor of atom $j$} \,\}$,
% the set of the first neighbors of the $j$-th atom by $\Omega_j$,
we have:
\begin{align}\label{eqn:split-rho-q2}
  \sum_j\sum_{k\in\Omega_j}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle
  \sum_j\sum_{k\notin\Omega_j}q_j q_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle.
We assume two charges are independent if they are not nearest neighbors.
Then
second part on the r.h.s. of the Eqn.~\eqref{eqn:split-rho-q2} is
approximately $\frac{N-\#(\Omega_j)}{N}\rho_q(\v r')\rho_q(\v r'')\approx
\rho_q(\v r')\rho_q(\v r'')$,
where $\#(\Omega_j)$ is the number of nearest neighbors considered.
The approximation is valid only when the number
of neighbors in $\Omega_j$ is small comparing with the number of charges
in the system.
% If the number of atoms in the system is
% large, this should be a good approximation.
Therefore, the first part
on the r.h.s. of Eqn.~\eqref{eqn:split-rho-q2} is actually the
pairwise charge correlation function:
\begin{align}\label{eqn:fnc-correlation-water}
  C_{q^2}(\v r', \v r'') \approx
  \sum_j\sum_{k\in\Omega_j}q_jq_k\delta(\v r' - \v r_j)\delta(\v r'' - \v r_k)
  \bigg\rangle.
% If connected by the chemical bond, the position of the first neighbors
% $k$ is not independent with the atom $j$, so their correlation is
% strong, and should be calculated explicitly to improve the quality of
% the error estimate.
By inserting Eqn.~\eqref{eqn:fnc-correlation-water} into
the correlation error, we have:
  \mathcal E_{\textrm{correlation}}(\v r_i)
  \approx q_i^2\sum_j\sum_{k\in\Omega_j}q_jq_k
  \v K(\v r_i , \v r_j)\cdot\v K(\v r_i , \v r_k)
  \big\rangle_{\v r_j,\v r_k}
The quality of this approximation depends on how to define the
``nearest neighbors''. In the system of liquid and gas, the
correlation is only short-ranged, so it is possible to systematically
improve the accuracy by considering more and more neighbors as long as
the number of neighbors is small comparing with the size of the system.

As an example, we consider a typical three point charge water model
TIP3P~\cite{jorgensen1983comparison}: each hydrogen atom carries a
charge of $q_{\textrm{H}} = +0.417e$ and each oxygen atom carries a
charge of $q_{\textrm{O}} = -0.834e$. Within one molecule, the H-O
bond and H-O-H angle are rigid, so the correlation within one molecule
is very strong.  Therefore, in this case, we define the nearest neighbor by the
atoms within the same water molecule.  Denoting the index set of the oxygen
atoms by $\Omega_{\textrm{O}}$, and the index set of the hydrogen
atoms by $\Omega_{\textrm{H}}$, we have:
  \mathcal E_{\textrm{correlation}}(\v r_i)
  q_i^2\sum_{j\in\Omega_{\textrm{O}}}\sum_{k\in\Omega_j}q_jq_k
  \v K(\v r_i , \v r_j)\cdot\v K(\v r_i , \v r_k)
  \big\rangle_{\v r_j,\v r_k} \\ \nonumber
  q_i^2\sum_{j\in\Omega_{\textrm{H}}}\sum_{k\in\Omega_j}q_jq_k
  \v K(\v r_i , \v r_j)\cdot\v K(\v r_i , \v r_k)
  \big\rangle_{\v r_j,\v r_k} \\\nonumber
  \v K(\v r_i , \v r')\cdot
  2\big\langle
  q_{\textrm{H}}\v K(\v r_i , \v r' + \v s_{\textrm{O}})
  \big\rangle\,
  \rho_{\textrm{O}}(\v r')
  \d d\v r' \\\label{eqn:c-error-1}
  & \,+
  \v K(\v r_i , \v r')\cdot
  q_{\textrm{O}}\v K(\v r_i , \v r' - \v s_{\textrm{O}})+
  q_{\textrm{H}}\v K(\v r_i , \v r' + \v s_{\textrm{H}})
  \big\rangle\,\rho_{\textrm{H}}(\v r')
  \d d\v r'
If positions of the three atom within one molecule are
$\v r_{\textrm{H}_1}$, $\v r_{\textrm{H}_2}$ and $\v r_{\textrm{O}}$, then
% $\v s_{\textrm{H}_1} = \v r_{\textrm{H}_1} - \v r_{\textrm{O}}$, 
% $\v s_{\textrm{H}_2} = \v r_{\textrm{H}_2} - \v r_{\textrm{O}}$,
% $\v s_{\textrm{O}} = \v r_{\textrm{O}} - \v r_{\textrm{H}_1}$ and
% $\v s_{\textrm{H}} = \v r_{\textrm{H}_2} - \v r_{\textrm{H}_1}$.
$\v s_{\textrm{O}} = \v r_{\textrm{H}_1} - \v r_{\textrm{O}}$ and
$\v s_{\textrm{H}} = \v r_{\textrm{H}_2} - \v r_{\textrm{H}_1}$.
The two ensemble averages in Eqn.~\eqref{eqn:c-error-1}
is taken over all possible water directions 
at position $\v r'$. As an approximation, we
assume these averages are  spatially uniform, namely independent with $\v r'$.
It may be  problematic
in the regions where the preference of water  direction is different
with others.
If the error force kernel
has the form $\v K(\v r_i , \v r_j) = \v K(\v r_i - \v r_j)$, 
the integrals can be written in the convolution form and accelerated by the FFT.
% In the example considered by the present paper in section~\ref{sec:tmp3},
% it is shown to be a good approximation to obtain the estimate of the
% correlation error.
One possible way to calculate the ensemble averages in
Eqn.~\eqref{eqn:c-error-1} is by the Fourier transform
of the error force kernel, for example
  q_{\textrm{H}}\v K(\v r_i - \v r' + \v s_{\textrm{O}})
  \frac1V\bigg\langle
  \sum_{\v m}
  q_{\textrm{H}}\hat{\v K}(\v m)
  e^{2\pi i\v m\cdot(\v r_i - \v r' + \v s_{\textrm{O}})} 
  \bigg\rangle \\\nonumber
  \frac1V\sum_{\v m}
  \langle
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{O}}}
  \rangle\,
  \hat{\v K}(\v m)\,
  e^{2\pi i\v m\cdot(\v r_i - \v r')},
where ``$\wedge$'' denotes the forward Fourier transform. By denoting
  \hat T_{\textrm{O}}(\v m)
  2\langle
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{O}}}
  \rangle,\\
  \hat T_{\textrm{H}}(\v m)
  \langle
  q_{\textrm{O}}\,e^{-2\pi i\v m\cdot\v s_{\textrm{O}}} +
  q_{\textrm{H}}\,e^{2\pi i\v m\cdot\v s_{\textrm{H}}}
  \rangle,
the Eqn.~\eqref{eqn:c-error-1} becomes:
  \mathcal E_{\textrm{correlation}}(\v r_i)
  =&\,
  \v K(\v r_i - \v r')\cdot
  (\hat T_{\textrm{O}}\hat{\v K})^{\vee}(\v r_i - \v r')
  \,\rho_{\textrm{O}}(\v r')
  \,\d d\v r' \\\label{eqn:c-error-water}
  & \,+
  \v K(\v r_i - \v r')\cdot
  (\hat T_{\textrm{H}}\hat{\v K})^{\vee}(\v r_i - \v r')
  \,\rho_{\textrm{H}}(\v r')
  \,\d d\v r',
where ``$\vee$'' denotes the backward Fourier transform.
In the Fourier space, the error force kernel $\v K(\v r)$ is
influenced by the prefactors $T_{\textrm{O}}$ and $T_{\textrm{H}}$,
which are accounting for the atomic correlation within one water
molecule.

@@

From the precise calculation of the reciprocal summation to the SPME
fast algorithm, the electrostatic interaction is approximated by two
steps.  The first step is the truncation of the reciprocal summation,
i.e. Eqn. \eqref{eqn:tmp6}. The second step is the approximation of
$e^{2\pi i\v m\cdot\v r}$ by B-spline
interpolation. It has been shown that the
error due to the first step is negligible comparing with the second
step,
at least for the parameters of practical interest~\cite{wang2010optimizing}.
Numerical
examples supporting this argument in inhomogeneous systems are also
given in Section \ref{sec:tmp3}.  Therefore, to estimate the force
error, it is enough to compare the SPME force with the
\emph{truncated} Ewald force \eqref{eqn:tmp8}.

As mentioned above, the SPME error mainly originates from approximating

@@

the analytical expression of $A(\v m, \v r)$~\cite{schoenberg1987cardinal}:

@@

The error force of the ik-differentiation is therefore

@@

estimates, and leave all details of the derivation 

@@

valleys of the high-wave-number terms may cancels in the the
ensemble-averaged error estimate.

@@

so it can be calculated once and stored in a table for future uses.

@@

\subsection{Error estimate of the analytical differentiation}
\label{sec:error-ana}
